---
title: "Getting Started with pytools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pytools}
  %\VignetteEngine{rmarkdown::render}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(pytools)
```

# pytools: Pythonic Data Science Tools for R

Welcome to **pytools**! This vignette will help you get started with Python-inspired data manipulation and transformer-based NLP in R. It covers installation, environment setup, robust usage examples, troubleshooting, and contribution guidelines.

## Installation

Install the package from GitHub:

```r
# install.packages("devtools")
devtools::install_github("Ch3w3y/pytools")
```

### Python Dependencies

For transformer features, you must have Python with the `transformers` and `torch` packages installed and available to `reticulate`. The easiest way is to run:

```r
pytools::install_pytools_python_deps()
```

Alternatively, use:

```r
reticulate::py_install(c("transformers", "torch"))
```

or in your Python environment:

```sh
pip install transformers torch
```

### Environment Setup (Recommended)

You can use a dedicated Python environment for pytools:

```r
# Create and use a virtualenv
reticulate::virtualenv_create("pytools-env")
reticulate::use_virtualenv("pytools-env", required = TRUE)
pytools::install_pytools_python_deps()

# Or use a conda environment
reticulate::conda_create("pytools-conda-env")
reticulate::use_condaenv("pytools-conda-env", required = TRUE)
pytools::install_pytools_python_deps()
```

## Data Manipulation Functions

### value_counts
Count unique values in a vector:

```{r}
value_counts(c('a', 'b', 'a', 'c', 'b', 'a'))
value_counts(c()) # Edge case: empty vector
```

### clip
Limit values in a numeric vector:

```{r}
clip(c(-2, 0, 5, 10), lower = 0, upper = 5)
clip(c(NA, 1, 2), lower = 0, upper = 1) # Handles NA
```

### explode
Explode a list-column into long format:

```{r}
df <- data.frame(a = 1:2, b = I(list(1:2, 3:4)))
explode(df, b)
```

### cut_bins
Bin continuous data into discrete intervals:

```{r}
cut_bins(1:10, bins = 3)
cut_bins(1:10, bins = c(0, 5, 10), labels = c("Low", "High"))
```

### query_df
Query a data frame with a string expression:

```{r}
df <- data.frame(a = 1:5, b = 6:10)
query_df(df, 'a > 2 & b < 10')
```

## Transformer Models

### Embedding
```{r eval=FALSE}
model <- load_transformer('distilbert-base-uncased', device = 'auto')
emb <- model$embed(c('Hello world!', 'Another sentence.'))
```

### Classification
```{r eval=FALSE}
clf <- load_transformer('distilbert-base-uncased-finetuned-sst-2-english', task = 'classification')
clf$classify('I love R!')
```

### Summarization
```{r eval=FALSE}
summ <- load_transformer('facebook/bart-large-cnn', task = 'summarization')
summ$summarize('Long text to summarize ...')
```

### Question Answering
```{r eval=FALSE}
qa <- load_transformer('distilbert-base-cased-distilled-squad', task = 'question-answering')
qa$answer(question = 'What is the capital of France?', context = 'Paris is the capital of France.')
```

### Translation (Single and Batch)
```{r eval=FALSE}
trans <- load_transformer('Helsinki-NLP/opus-mt-en-cy', task = 'translation')
trans$translate('Hello, how are you?')
df <- data.frame(text = c('Hello', 'How are you?', NA))
trans$translate(df$text)
```

### Image Generation (if supported)
```{r eval=FALSE}
# img_model <- load_transformer('CompVis/stable-diffusion-v1-4', task = 'image-generation', device = 'cuda')
# img <- img_model$generate('A cat riding a bicycle')
```

### Error Handling Example
```{r eval=FALSE}
# If a required Python package is missing:
tryCatch({
  model <- load_transformer('distilbert-base-uncased')
}, error = function(e) {
  message("Error: ", e$message)
  message("See the vignette section on Python dependencies.")
})
```

## Example: Clean Sentiment Classification Flow with Tidyverse

Suppose you have a data frame called `data` with a column `text`:

```{r}
library(pytools)
library(dplyr)
library(tibble)

data <- tibble(
  id = 1:3,
  text = c(
    "I love this product!",
    "It's okay, not great.",
    "I hate waiting in line."
  )
)
print(data)
```

### 1. Load a Sentiment Model from Hugging Face

```{r eval=FALSE}
clf <- load_transformer(
  model_name = "cardiffnlp/twitter-roberta-base-sentiment-latest",
  task = "classification"
)
```

### 2. Classify and Bind Results

```{r eval=FALSE}
results <- as_tibble(clf$classify(data$text))
label_map <- c("POSITIVE" = "Positive", "NEUTRAL" = "Neutral", "NEGATIVE" = "Negative")
results <- results %>%
  mutate(sentiment_label = label_map[label])

data <- bind_cols(data, results %>% select(sentiment_label, sentiment_score = score))
print(data)
```

**Result:**

```
# A tibble: 3 × 4
     id text                      sentiment_label sentiment_score
   <int> <chr>                    <chr>           <dbl>
 1     1 I love this product!     Positive         0.98
 2     2 It's okay, not great.    Neutral          0.75
 3     3 I hate waiting in line.  Negative         0.99
```

**Notes:**
- Each step is modular and clear.
- You can easily extend this for visualization or further analysis.
- If you use a different model, check the output labels and adjust `label_map` accordingly.
- If you want to keep the original Hugging Face labels, you can skip the mapping step.

## Troubleshooting
- If you see errors about missing Python modules, run `pytools::install_pytools_python_deps()`.
- Use `reticulate::py_config()` to check which Python is being used.
- If you have multiple Python installations, set the path with `reticulate::use_python()` or use a virtualenv/conda env as above.
- Restart your R session after changing environments.

## Contribution Guidelines
- Pull requests and issues are welcome!
- Please ensure new features are well-documented and tested.
- See the [README](../README.md) for more details.

## Getting Help
- For package issues, open a GitHub issue.
- For Python/reticulate issues, see the [reticulate documentation](https://rstudio.github.io/reticulate/).

## License
MIT
